Nix for Polyglot, Reproducible Data Science Workflows

### 2025-10-24 paper.qmd
better conclusion

@@

While [rix]{.pkg} and [rixpress]{.pkg} (and [ryxpress]{.pkg}) provide
significant steps forward, it is important to acknowledge current limitations.
[rixpress]{.pkg} (and [ryxpress]{.pkg}) in particular requires further
development, especially to enhance the ease of debugging complex pipelines. Its
visualisation capabilities are also still fairly limited; for instance, they do
not currently indicate outdated derivations when a user modifies code that would
necessitate a rebuild. These areas represent important avenues for future work
to further mature the framework and improve the user experience.

>>>>

While the reliance on the `rstats-on-nix` fork might initially raise
sustainability concerns, this risk should be contextualized. The fork serves a
focused purpose: it updates CRAN and [Julia]{.proglang} package registries daily
and backports fixes when builds break. Importantly, users are never locked into
this fork—they retain the ability to point directly to any commit in the
upstream `nixpkgs` repository, maintaining full access to the official
[Nix]{.pkg} ecosystem. The fork introduces no fundamental architectural changes
or proprietary extensions; it functions as a convenience layer providing tested
package sets and dated snapshots. Should maintenance cease (which is unlikely,
as I directly use it myself), users could seamlessly transition to upstream
`nixpkgs` commits, though with trade-offs: less frequent [R]{.proglang} and
[Julia]{.proglang} package updates. The fork's value lies not in creating
dependency, but in reducing friction through testing infrastructure and curated
"known-good" dates. Thus, the fork represents an enhancement to user experience
rather than a critical single point of failure—the underlying [Nix]{.pkg}
infrastructure and official `nixpkgs` repository remain the foundation, ensuring
long-term viability of environments created with [rix]{.pkg}.

@@

Adopting [Nix]{.pkg}-based workflows represents a significant paradigm shift, and
researchers should approach it incrementally. A reasonable pathway begins with
using [rix]{.pkg} to generate reproducible environments for new projects, while
maintaining existing workflows for ongoing work. This allows learning [Nix]{.pkg}
concepts without risking active research. As comfort grows, researchers can
experiment with [rixpress]{.pkg} for smaller analyses before tackling complex
polyglot pipelines. Institutions can support adoption by providing workshops,
maintaining institutional binary caches with private packages, and designating
local experts. The investment pays increasing returns: early projects require
substantial learning, but subsequent projects leverage accumulated knowledge and
reusable environment definitions. For research groups, having one member develop
[Nix]{.pkg} expertise can enable the entire team to benefit from reproducible
environments without everyone climbing the learning curve. Ultimately, the
decision to adopt these tools depends on project timescales, reproducibility
requirements, and institutional context—but the framework exists for those who
need it.

While [rix]{.pkg} and [rixpress]{.pkg} (and [ryxpress]{.pkg}) provide
significant steps forward, it is important to acknowledge current limitations.
Debugging complex [rixpress]{.pkg} pipelines remains more difficult than
debugging [targets]{.pkg} workflows, particularly when failures occur deep in
the [Nix]{.pkg} build process. The visualization capabilities, while useful, lack
features like automatic detection of outdated derivations that make tools like
[targets]{.pkg} particularly developer-friendly. Cross-platform reproducibility,
while generally strong, has known limitations on macOS due to system framework
dependencies. These are not insurmountable problems—they represent areas for
continued development and community contribution. Researchers adopting these
tools should do so with realistic expectations: they provide powerful
capabilities for deep reproducibility and polyglot workflows, but require
patience during the learning phase and may occasionally present challenges that
simpler tools avoid.
